Assumptions of Physics blueprint: information granularity as a common foundation for geometry, measures and probability

### 2021-04-09 Overviews/Informational granularity/InformationGranularity.tex
InfGran: biblio fix
InfGran: Mark's review

Note that on Page 21 de Finetti\cite{definetti} specifically says that $P(E) \geq P(E’)$ implies $E \geq E’$ but not vice-versa. And that we can have $P(A)=P(B)=0$ even if $A<B$ if $A$ is impossible and $B$ is not. So, the ordering he has in mind is fineness for us. And then he says that this means that the 4 postulates give a non-Archimedean structure. But you can make it Archimedean by ignoring the probabilities that are “infinitesimally small”: those that multiplied by a number, however big, will never become the certainty. On the other hand, Villegas\cite{villegas} starts already with an ordering where the infinitesimals are equal to the empty set. This maps to $\lequ$ defined later, where the comparison is done up to infinitesimals.

@@

It would be nice to be able to prove the following either true or false.

\begin{conj}\label{uniformconsequence}
    Let $\tdomain$ be uniform. Let $\stmt, \stmt_1 \in \tdomain$ such that $\stmt \sfiner \stmt_1$. Then there exist $\stmt_2 \in \tdomain$ such that $\stmt_2 \narrower \stmt_1$ and $\stmt \eqgran \stmt_2$.
\end{conj}

[TODO: clean-up what follows]

The start of a proof....
Let $S = \{\stmt[t]\in\tdomain : \stmt[t]\narrower\stmt_1 \text{ and } \stmt[t]\finer\stmt\}$. We will find a maximal element with respect to fineness (OR narrowness??).

The difficulty here is that for Zorn we need to find an upper bound of an arbitrary chain which also lies in $S$. But such an upper bound is entirely unclear on how to obtain without using the proposition to begin with. Uniformity is difficult to make much use of for infinite domains (beyond that $S$ is nonempty).

This set $S$ is nonempty, since any possibility narrower than $\stmt_1$ must also be finer than $\stmt$ by uniformity, and the statement $\stmt_1$ must not be the impossibility since it is strictly coarser than the statement $\stmt$.

Now, given any chain (with respect to narrowness), consider its disjunction. But again this may be coarser than $\stmt$!

Note: problem with chains for fineness. Taking the disjunction of a chain with respect to fineness can yield something much coarser (i.e. outside of $\{\stmt[t]\in\tdomain : \stmt[t]\narrower\stmt_1 \text{ and } \stmt[t]\finer\stmt\}$) if they do not overlap.

Idea: given chains in fineness, can we construct new chains which are, at each level, equigranular, but now they are also chains in narrowness?

Lemma: If $C$ is a chain in fineness and $\stmt$ is such that $\stmt[t]\finer\stmt$ for all $\stmt[t]\in C$, then there exists a set $N$ and a bijection $f:C\to N$ such that $N$ is a chain in narrowness, $\OR N \narrower \stmt$, and for all $\stmt[t]\in C$, $\stmt[t]\eqgran f(\stmt[t])$. 
Use this to move chains in fineness over to chains in narrowness and then can find maximal elements in the set $S$ above ``up to equigranular equivalence classes".

Problem: this is actually much harder. Prop 6 is the case of this Lemma for a chain of length 2 and the top of the chain being equal to $\stmt$.

@@

\begin{conj}
	Equivalence at $\stmt[u]$ scale is an equivalence class.
\end{conj}

@@

	If we assume a uniform domain, we would have totality and we may also be able to show that incompatible infinitesimals can always be found. It may be that totality is not needed for proposition \ref{infinitesimal_disjunction}. It may be that we need to amend the definition of infinitesimals, such that the disjunction of two infinitesimals is an infinitesimals. We could also force the transitivity, even if the disjunction of two infinitesimals is not an infinitesimals. This would be like saying that, yes, in principles the assertions correspond to different granularity levels, but we are not to distinguish between them because the different is effectively zero. Understanding which options is best requires more work.
\end{remark}

\begin{conj}
	Let $\stmt_1, \stmt_2 \in \tdomain$ such that $\stmt_1 \equ \stmt_2$ and $\stmt_1 \finer \stmt_2$, then for all $\stmt_1 \finer \stmt \finer \stmt_2$ we have $\stmt_1 \equ \stmt \equ \stmt_2$.
\end{conj}

@@

[TODO: is there anything to be salvaged between here]

We start with a simpler case:

\begin{prop}
	Let $\stmt_1,\stmt,\stmt[v]_1\in\tdomain$ such that $\stmt_1\finer\stmt\finer\stmt_1\OR\stmt[v]_1$, and suppose $\stmt[v]_1$ is infinitesimal with respect to $\stmt[u]$. Then $\stmt\equ\stmt_1$. 
\end{prop}
\begin{proof}[Using Prop \ref{uniformconsequence}]
	If both of the $\finer$'s are actually equigranular, there is nothing to show. Next consider the case that $\stmt_1\sfiner\stmt\finer\stmt_1\OR\stmt[v]_1$. Then by Proposition \ref{uniformconsequence} there exists $\stmt_2$ such that $\stmt_2\narrower\stmt$ and $\stmt_2\eqgran\stmt_1$. But how do we put this together?? What about the even simpler version of the lemma with $\narrower$ instead of $\finer$? Just use $\stmt\OR (\stmt_1\OR\stmt[v]_1 \AND \NOT\stmt)$.
\end{proof}

This seems problematic already:

We have $\stmt_1\finer\stmt\finer\stmt_1\OR\stmt[v]_1$. We need to find some $\stmt[w],\stmt[w]_1$ infinitesimal with respect to $\stmt[u]$, such that $\stmt\OR\stmt[w]\eqgran\stmt_1\OR\stmt[w]_1$. Since we don't know anything about the overlap of $\stmt$ and $\stmt_1$, we can't use any operations like $\stmt\AND\NOT\stmt_1$ to find something small. We need to know that additional statements exist in order to find $\stmt[w]$ and $\stmt[w]_1$.

For example, what if every infinitesimal statement was narrower $(\narrower)$ than $\stmt$? In this case, we could have something like $\stmt_1\sfiner\stmt\sfiner\stmt_1\OR\stmt[x]$ for all statements (including all infinitesimals) $\stmt[x]$. It would then not be the case that $\stmt\equ\stmt_1$.

\begin{prop}[Need something like this but not clear if correct!]
	Suppose $\stmt_1\equ\stmt_2$ and $\stmt_2\equ\stmt_3$. Then there exist statements  $\stmt[v]_1,\stmt[v]_2,\stmt[w]_2,\stmt[w]_3$, infinitesimal in $\stmt[u]$, such that:
	$$
	\stmt_1\OR\stmt[v]_1\OR\stmt[w]_2 \eqgran \stmt_2\OR\stmt[v]_2\OR\stmt[w]_2 \eqgran \stmt_3\OR\stmt[w]_3\OR\stmt[v]_2.
	$$
\end{prop}

Note that the existence of the right combination of incompatible statements is sufficient. However, that is stronger than we need and may be harder to prove. Problem: we have very few propositions about existence of statements! It is hard to ensure certain ones exist.

Consider the following (extreme) case. Suppose $\stmt_1\equ\stmt_2$ and $\stmt_2\equ\stmt_3$, but the infinitesimals which realize these relationships are $\stmt[v]_1$ and $\stmt[v]_2$, as follows:
$$
\stmt_1\OR\stmt[v]_1 \eqgran \stmt_2\OR\stmt[v]_2,\ \stmt_2\OR\stmt[v]_1\eqgran\stmt_3\OR\stmt[v]_2
$$
Without further axioms, it may not be possible to put this together to get transitivity (i.e. to find that $\stmt_1\equ\stmt_3$).

\begin{prop}
	The relation $\equ$ is an equivalence relation.
\end{prop}
\begin{proof}
	Reflexivity and symmetry are immediate since $\eqgran$ is an equivalence relation. For transitivity, suppose $\stmt_1\equ\stmt_2$ and $\stmt_2\equ\stmt_3$. Let $\stmt[v]_1,\stmt[v]_2,\stmt[w]_2,\stmt[w]_3$ be infinitesimals with respect to $\stmt[u]$, such that $\stmt_1\OR\stmt[v]_1\eqgran\stmt_2\OR\stmt[v]_2$ and $\stmt_2\OR\stmt[w]_2\eqgran\stmt_3\OR\stmt[w]_3$. Then we have:
	$$
	\stmt_1\OR\stmt[v]_1\OR\stmt[w]_2 \eqgran \stmt_2\OR\stmt[v]_2\OR\stmt[w]_2 \eqgran \stmt_3\OR\stmt[w]_3\OR\stmt[v]_2.
	$$
	Since $\stmt[v]_1\OR\stmt[w]_2$ and $\stmt[w]_3\OR\stmt[v]_2$ are infinitesimal with respect to $\stmt[u]$ (being finite disjuncts of infinitesimals), we have $\stmt_1\equ\stmt_3$. 
\end{proof}

[TODO: and here?]

@@

@@

\begin{conj}
	The two definitions are equivalent.
\end{conj}

@@

[TODO: would remove this]

{\bf Warning:} Transitivity doesn't seem to work as stated (see the proof attempt below). We propose the following modification: say $\stmt_1\lequ\stmt_2$ if there exist statements $\stmt_1'$ and $\stmt_2'$ with $\stmt_1'\equ\stmt_1$ and $\stmt_2'\equ\stmt_2$ such that $\stmt_1'\finer\stmt_2'$. Note that {\bf by totality}, it is sufficient to just say $\stmt_1\finer\stmt_2'$ (no need for $\stmt_1'$).

[TODO: proof not clear]

\begin{prop}
	$\stmt_1 \leu \stmt_2$ if for every statement $\stmt[v]$ infinitesimal in $\stmt[u]$, we have $\stmt_1\OR\stmt[v] \sfiner \stmt_2$. 
\end{prop}
\begin{proof}
	Since $\impossibility$ is infinitesimal, we have $\stmt_1 = \stmt_1\OR\impossibility \sfiner \stmt_2$, and so by Definition \ref{smalleru}, we have that $\stmt_1\lequ\stmt_2$. Now, we must show it is not the case that $\stmt_1\equ \stmt_2$.

	Let $\stmt[v]_1,\stmt[v]_2\in\tdomain$ be infinitesimals. Then $\stmt_1\OR\stmt[v]_1 \sfiner \stmt_2 \finer\stmt_2\OR\stmt[v]_2$. We conclude that it cannot be the case that $\stmt_1 \OR \stmt[v]_1 \eqgran \stmt_2 \OR \stmt[v]_2$ for any infinitesimals (with respect to $\stmt[u]$) $\stmt[v]_1,\stmt[v]_2$. 
\end{proof}

We next show that $\lequ$ is a pre-order, and that it is a partial order with respect to $\equ$-equivalence classes.

@@

[TODO: not clear that this works... don't we have to find infinitesimals that do not overlap?]
\begin{proof}
	Reflexivity is immediate since for any $\stmt\in\tdomain$, we have $\stmt\finer\stmt$, so $\stmt\lequ\stmt$. (This works for both the new and old definition)

	Transitivity requires a little more work. Let $\stmt_1,\stmt_2,\stmt_3\in\tdomain$ with $\stmt_1\lequ\stmt_2\lequ\stmt_3$.

	{\bf Using the original definition:}
	We have several cases to consider.

	\emph{Case 1:} $\stmt_1\finer\stmt_2\finer\stmt_3$. Then $\stmt_1\finer\stmt_3$ and so $\stmt_1\lequ\stmt_3$.

	\emph{Case 2:} $\stmt_1\equ\stmt_2\equ\stmt_3$. Then $\stmt_1\equ\stmt_3$ (since $\equ$ is an equivalence relation), and so $\stmt_1\lequ\stmt_3$.

	\emph{Case 3:} $\stmt_1\equ\stmt_2\finer\stmt_3$. Then for some infinitesimals $\stmt[v]_1$ and $\stmt[v]_2$ (with respect to $\stmt[u]$), we have:
	$$
	\stmt_1\finer\stmt_1\OR\stmt[v]_1\eqgran\stmt_2\OR\stmt[v]_2\finer\stmt_3\OR\stmt[v]_2
	$$
	Unclear what to do here -- need $\stmt_3$ alone on the right-hand side somehow. Case 4 has the same problem. Let's try it with the modified definition (in the ``warning").

	{\bf Using the new definition:}
	Let $\stmt_1'\equ\stmt_1$ and $\stmt_2'\equ\stmt_2$ with $\stmt_1'\finer\stmt_2'$, and let $\stmt_2''\equ\stmt_2$ and $\stmt_3''\equ\stmt_3$ with $\stmt_2''\finer\stmt_3''$. 
	Then for an appropriate choice of infinitesimals (with respect to $\stmt[u]$) $\stmt[v]_i,\stmt[w]_i,\stmt[v]_i',\stmt[w]_i'$, $i=1,2,3$, we have:
	$$
	\stmt_1\OR\stmt[v]_1\OR\stmt[w]_2'\OR\stmt[w]_3'' \eqgran \stmt_1'\OR\stmt[v]_1'\OR\stmt[w]_2'\OR\stmt[w]_3'' \finer \stmt_2'\OR\stmt[v]_1'\OR\stmt[w]_2'\OR\stmt[w]_3'' \eqgran \stmt_2''\OR\stmt[w]_2''\OR\stmt[v]_1'\OR\stmt[w]_3'' \finer
	$$
	$$
	\stmt_3''\OR\stmt[w]_2''\OR\stmt[v]_1'\OR\stmt[w]_3'' \eqgran
	\stmt_3\OR\stmt[w]_3 \OR\stmt[w]_2''\OR\stmt[v]_1'
	$$
	Since $\stmt_1 \equ \stmt_1\OR\stmt[v]_1\OR\stmt[w]_2'\OR\stmt[w]_3''$ and $\stmt_3 \equ \stmt_3\OR\stmt[w]_3 \OR\stmt[w]_2''\OR\stmt[v]_1' $, we have that $\stmt_1\lequ\stmt_3$.

@@

\begin{remark}
	content...
\end{remark}

@@

\subsubsection{Collapsing Orders}
[TODO: didn't we find a reference for this?]

Searching around did not yield the theorems desired, so we will discuss the tools needed here. It turns out that it is not so straightforward to collapse a poset by an equivalence relation. Let's start with a possible definition:

If $(P,\leq)$ is a poset and $\sim$ is a collapsing equivalence relation, then endow the set $P/\sim$ with the relation $\leq/\sim$ defined by $x/\sim \leq/\sim y/\sim$ whenever there exist $x'\in x/\sim$ and $y'\in y/\sim$ with $x'\leq y'$.

It turns out that this is not sufficient. Here is an important example. Let our set be $\{w,x,y,z\}$ and a poset relation defined by $x\leq w$, $y\leq w$, $y\leq z$, with the following Hasse diagram:

@@

Note that neither of the above problems are possible with {\bf totality}. It seems that what makes the most sense is collapsing by some equivalence relation, and then extending both the ordering and the equivalence relation to generate a legitimate poset. We can explore this later if needed.

@@

\bibliographystyle{apsrev4-2}
\bibliography{bibliography}

>>>>

Note that on Page 21 de Finetti\cite{definetti} specifically says that $P(E) \geq P(E')$ implies $E \geq E'$ but not vice-versa. And that we can have $P(A)=P(B)=0$ even if $A<B$ if $A$ is impossible and $B$ is not. So, the ordering he has in mind is fineness for us. And then he says that this means that the 4 postulates give a non-Archimedean structure. But you can make it Archimedean by ignoring the probabilities that are ``infinitesimally small'': those that multiplied by a number, however big, will never become the certainty. On the other hand, Villegas\cite{villegas} starts already with an ordering where the infinitesimals are equal to the empty set. This maps to $\lequ$ defined later, where the comparison is done up to infinitesimals.

@@

We expect the following to not be provable from the given axioms:

\begin{desid}\label{uniformconsequence}
    Let $\tdomain$ be uniform. Let $\stmt, \stmt_1 \in \tdomain$ such that $\stmt \sfiner \stmt_1$. Then there exist $\stmt_2 \in \tdomain$ such that $\stmt_2 \narrower \stmt_1$ and $\stmt \eqgran \stmt_2$.
\end{desid}
However, it would be nice to understand why as it would give us insights into which additional premises are needed. The idea is that we can always ``move" a statement to an equigranular one to sit within any other (coarser) statement. This would allow us to make use of various properties of narrowness. Interestingly, this would not give additional insight into limiting behaviors, as equigranular sequences may have vastly different limits (and so ``moving" a sequence could drastically change the equigranularity class of the limit).

@@

When the definitions and axioms are solidified, the following will be essential:

\begin{desid}
	Equivalence at $\stmt[u]$ scale is an equivalence relation.
\end{desid}

@@

	If we assume a uniform domain, we would have totality and we may also be able to show that incompatible infinitesimals can always be found. It may be that totality is not needed for proposition \ref{infinitesimal_disjunction}. It may be that we need to amend the definition of infinitesimals, such that the disjunction of two infinitesimals is an infinitesimals. We could also force the transitivity, even if the disjunction of two infinitesimals is not an infinitesimals. This would be like saying that, yes, in principles the assertions correspond to different granularity levels, but we are not to distinguish between them because the different is effectively zero. Understanding which options are best requires more work.
\end{remark}

Another essential property of equivalence at $\stmt[u]$ scale is that it collapses the fineness ordering, formalized by the following statement:

\begin{desid}
	Let $\stmt_1, \stmt_2 \in \tdomain$ such that $\stmt_1 \equ \stmt_2$ and $\stmt_1 \finer \stmt_2$. Then for all $\stmt_1 \finer \stmt \finer \stmt_2$ we have $\stmt_1 \equ \stmt \equ \stmt_2$.
\end{desid}

@@

A special case of the above may be a good first goal for testing appropriate axioms, or perhaps it could end up being an axiom itself:

\begin{desid}
	Let $\stmt_1,\stmt,\stmt[v]_1\in\tdomain$ such that $\stmt_1\finer\stmt\finer\stmt_1\OR\stmt[v]_1$, and suppose $\stmt[v]_1$ is infinitesimal with respect to $\stmt[u]$. Then $\stmt\equ\stmt_1$. 
\end{desid}

In addition to defining equivalence at scale $\stmt[u]$, we also wish to have the ordering. Eventually, the goal is to use this as a qualitative probability, as in Fishburn \cite{fishburnsurvey}.

@@

This is the final notion of ordering. After narrowness orders the actual content of the statements and fineness gives a very precise ordering of size, ordering by scale in $\stmt[u]$ aims to give a less universal notion of size relative to a fixed unit statement $\stmt[u]$, where we collapse by equivalence at scale.

An alternative definition is as follows.

@@

We still need to show the following:

\begin{desid}
	The two definitions are equivalent.
\end{desid}

@@

\begin{prop}
	If for every statement $\stmt[v]$ infinitesimal in $\stmt[u]$, we have $\stmt_1\OR\stmt[v] \sfiner \stmt_2$, then we have $\stmt_1 \leu \stmt_2$. 
\end{prop}
\begin{proof}
	First, using $\stmt[v] = \impossibility$ (an infinitesimal) in the hypothesis of the statement, we have $\stmt_1 = \stmt_1\OR\impossibility \sfiner \stmt_2$, and so by Definition \ref{smalleru}, we have that $\stmt_1\lequ\stmt_2$. Now, we must show it is not the case that $\stmt_1\equ \stmt_2$.

	Suppose that $\stmt_1 \equ \stmt_2$. By Definition \ref{smalleru}, let $\stmt[v]_1,\stmt[v]_2\in\tdomain$ be infinitesimals such that $\stmt_1 \OR \stmt[v]_1 \eqgran \stmt_2 \OR \stmt[v]_2$. But by assumption, $\stmt_1\OR\stmt[v]_1 \sfiner \stmt_2 \finer\stmt_2\OR\stmt[v]_2$. This contradicts the earlier equigranularity, and so it cannot be the case that $\stmt_1\equ\stmt_2$. 
\end{proof}

We need it to be the case that $\lequ$ is a pre-order, and that it is a partial order with respect to $\equ$-equivalence classes.

@@

Fortunately, the current definitions of $\lequ$ and $\equ$ behave well together:

@@

@@

\subsubsection{Collapsing Orders}
In Section 2.1 of \cite{quotientposets}, the idea of collapsing a poset by an equivalence relation is introduced.

If $(P,\leq)$ is a poset and $\sim$ is a collapsing equivalence relation, we may endow the set $P_{/\sim}$ with the relation $\leq_{/\sim}$ defined by $x_{/\sim} \leq_{/\sim} y_{/\sim}$ whenever there exist $x'\in x_{/\sim}$ and $y'\in y_{/\sim}$ with $x'\leq y'$ (this is the definition used in \cite{quotientposets}).

It turns out that $(P_{/\sim},\leq_{/\sim})$ is not always a poset. Here is an important example. Let our set be $\{w,x,y,z\}$ and a poset relation defined by $x\leq w$, $y\leq w$, $y\leq z$, with the following Hasse diagram:

@@

Lemma 2.1.3 in \cite{quotientposets} gives sufficient conditions on the interaction of $\leq$ and $\sim$ on $P$ for $(P_{/\sim},\leq_{/\sim})$ to be a poset. These essentially prevent problems of the above form from happening.

\begin{prop}
Suppose $(P,\leq)$ has a minimum element $m$ and that the equivalence class $m_{/\sim}$ contains no other elements. Further, suppose that if $x_{/\sim} \leq_{/\sim} y_{/\sim}$, then for all $x'\in x_{/\sim}$ there exists a $y'\in y_{/\sim}$ with $x'\leq y'$. 
Under these assumptions, $(P_{/\sim},\leq_{/\sim})$ is also a poset.
\end{prop}

As for a minimum element, we have the impossibility $\impossibility$ which will be unique in its equigranularity class, but being infinitesimal it will share its $\equ$ equivalence class, so perhaps modification of the above lemma will be necessary.

@@

\bibliography{bibliography}
